On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-09-02 paper.tex
Minor wording edits

The notion of a \emph{random} segment of a given length is precisely defined as follows. By symmetry, one endpoint of the segment can be assumed to lie in a fixed, reference tile. The position of this endpoint is \emph{uniformly} distributed on the tile. The segment orientation has a \emph{uniform} distribution on the set of all possible directions, and is \emph{independent} of the endpoint position. Solving the problem of how many tiles the segment visits on average, as will be seen, also answers the inverse question (segment length to visit a given number of tiles on average). A natural, related question is with that probability the segment visits the maximum number of tiles.

@@

\begin{proposition}
\label{prop: min suff set, form}
Given $a, b \in \mathbb R^+$ and $\tiles \in \mathbb N$, $\tiles \geq 3$, a minimal sufficient set $\mss = \{(i_3,j_3), (i_4.j_4), \ldots\}$ can be obtained as
\begin{align}
\label{eq: min suff set, form, i}
i_\tiles &= \left\lfloor \frac {(2\tiles-1)b^2+5a^2}{2(a^2+b^2)} \right\rfloor, \\
\label{eq: min suff set, form, j}
j_\tiles &= \left\lceil \frac {2\tiles a^2 + 3(b^2-a^2)}{2(a^2+b^2)} \right\rceil,
\end{align}
where $i_\tiles + j_\tiles-1 = \tiles$. All pairs $(i_\tiles,j_\tiles)$ are strictly below the line
\begin{equation}
\label{eq: upper bound, line}
j = \frac{i a^2}{b^2} - \frac{3a^2}{2b^2} + \frac 5 2,
\end{equation}
and above or on the line
\begin{equation}
\label{eq: lower bound, line}
j = \frac{i a^2}{b^2} - \frac{5a^2}{2b^2} + \frac 3 2.
\end{equation}
\end{proposition}

@@

For $a=b=1$ the set $\mss$ consists of points of the form $(i,i)$ and $(i,i-1)$, as is easily seen from Proposition~\ref{prop: min suff set, form}, and as illustrated in Figure~\ref{fig: pairs_bounds_1}. By symmetry, replacing each point $(i,i-1)$ by $(i-1,i)$ gives a set $\mss'$ which is also minimal sufficient. For this new set, the lower bounding line \eqref{eq: lower bound, line} can be replaced by the simpler $j=i$. The same approach used in the proof of Theorem~\ref{theo: funt, form} can be followed, but using this line. Thus $(\isolr, \jsolr)$ is obtained from
\begin{align}
(\isolr-2)^2 + (\jsolr-2)^2  &= \len^2, \\
\jsolr &= \isoli,
\end{align}
which gives
\begin{equation}
\label{eq: theo: funt, sq, form; proof 1}
\isolr = \len/\sqrt{2} + 2.
\end{equation}
The pair $(\isoli,\jsoli)$ resulting from \eqref{eq: theo: funt, sq, form; proof 1} is in $\mss'$, and is given in \eqref{eq: theo: funt, sq, form; i} and \eqref{eq: theo: funt, sq, form; j}. This establishes \eqref{eq: theo: funt, sq, form; funt, with i, j}.

***Alex
***Derivation of \eqref{eq: theo: funt, sq, form; funt, simpler} \ldots
***It looks like we may have to use different techniques for deriving \eqref{eq: theo: funt, sq, form; funt, with i, j} and \eqref{eq: theo: funt, sq, form; funt, simpler}. I don't see a way to use a common proof method. And \eqref{eq: theo: funt, sq, form; funt, with i, j} is needed because of $\isoli$, $\jsoli$, which are used in (the text below in preparation for) \S\ref{part: probmax: unit square grid, real lengths}.
\end{proof}

@@

\begin{proof}
***New proof without using the proposition (which has been moved after the theorem)
\end{proof}

In view of Theorem~\ref{theo: funta, form}, the average number of visited tiles has a remarkably simple dependence on the segment length, namely an affine function. Conversely, for any $\tiles>1$ it is immediate to compute the length of a random segment that visits $\tiles$ tiles on average, given as $\funta^{-1}(\tiles)$.

>>>>

To address the remaining two questions, the notion of a \emph{random} segment of a given length needs to be precisely defined. This is done as follows. By symmetry, one endpoint of the segment can be assumed to lie in a fixed, reference tile. The position of this endpoint is \emph{uniformly} distributed on the tile. The segment orientation has a \emph{uniform} distribution on the set of all possible directions, and is \emph{independent} of the endpoint position. Solving the problem of how many tiles the segment visits on average, as will be seen, also answers the inverse question (segment length to visit a given number of tiles on average). A natural, related question is with that probability the segment visits the maximum number of tiles.

@@

\begin{proposition}
\label{prop: min suff set, form}
Given $a, b \in \mathbb R^+$ and $\tiles \in \mathbb N$, $\tiles \geq 3$, a minimal sufficient set $\mss = \{(i_3,j_3), (i_4.j_4), \ldots\}$ can be obtained as
\begin{align}
\label{eq: min suff set, form, i}
i_\tiles &= \left\lfloor \frac {(2\tiles-1)b^2+5a^2}{2(a^2+b^2)} \right\rfloor, \\
\label{eq: min suff set, form, j}
j_\tiles &= \left\lceil \frac {(2\tiles-3) a^2 + 3b^2}{2(a^2+b^2)} \right\rceil,
% ***change expressions to the following to match the statement of Theorem~\ref{theo: funl, form}:
%i_\tiles &= \left\lfloor \frac{b^2 (\tiles-3)}{a^2+b^2} + \frac 5 2 \right\rfloor \\
%j_\tiles &= \left\lceil  \frac{a^2 (\tiles-3)}{a^2+b^2} + \frac 3 2 \right\rceil
\end{align}
where $i_\tiles + j_\tiles-1 = \tiles$. All pairs $(i_\tiles,j_\tiles)$ are strictly below the line
\begin{equation}
\label{eq: upper bound, line}
j = \frac{i a^2}{b^2} - \frac{3a^2}{2b^2} + \frac 5 2,
\end{equation}
and above or on the line
\begin{equation}
\label{eq: lower bound, line}
j = \frac{i a^2}{b^2} - \frac{5a^2}{2b^2} + \frac 3 2.
\end{equation}
\end{proposition}

@@

For $a=b=1$, the set $\mss$ consists of points of the form $(i,i)$ and $(i,i-1)$, as is easily seen from Proposition~\ref{prop: min suff set, form}, and as illustrated in Figure~\ref{fig: pairs_bounds_1}. By symmetry, replacing each point $(i,i-1)$ by $(i-1,i)$ gives a set $\mss'$ which is also minimal sufficient. For this new set, the lower bounding line \eqref{eq: lower bound, line} can be replaced by the simpler $j=i$. The same approach used in the proof of Theorem~\ref{theo: funt, form} can be followed, but using this line. Thus $(\isolr, \jsolr)$ is obtained from
\begin{align}
(\isolr-2)^2 + (\jsolr-2)^2  &= \len^2, \\
\jsolr &= \isoli,
\end{align}
which gives
\begin{equation}
\label{eq: theo: funt, sq, form; proof 1}
\isolr = \len/\sqrt{2} + 2.
\end{equation}
The pair $(\isoli,\jsoli)$ resulting from \eqref{eq: theo: funt, sq, form; proof 1} is in $\mss'$, and is given in \eqref{eq: theo: funt, sq, form; i} and \eqref{eq: theo: funt, sq, form; j}. This establishes \eqref{eq: theo: funt, sq, form; funt, with i, j}.

To show \eqref{eq: theo: funt, sq, form; funt, simpler}, it is first noted that for $\tiles \geq 3$ Corollary~\ref{theo: funl, sq, form} gives
\begin{equation}
\funl(\tiles) = \sqrt{\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil}.
\end{equation}
According to \eqref{eq: funt funl}, the function $\funt(\len)$ is given by the largest positive integer $t$ such that 
\begin{equation}
\label{eq: theo: funt, sq, form; proof 2}
\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil < \len ^ 2.
\end{equation}
Since the left-hand side of \eqref{eq: theo: funt, sq, form; proof 2} is an integer, the condition of being strictly less than $\len^2$ is equivalent to
\begin{equation}
\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil  \leq \left\lceil \len^2 \right\rceil  - 1,
\end{equation}
which in turn is the same as
\begin{equation}
\frac{(\tiles-3)^2} {2}\leq  \left\lceil \len^2 \right\rceil  - 1. 
\end{equation}
Finally, solving for $\tiles$ gives
\begin{equation}
\label{eq: theo: funt, sq, form; proof 3}
\tiles \leq \sqrt{2 \left\lceil \len^2 \right\rceil -2 } + 3.
\end{equation}
The desired quantity $\funt(\len)$, that is the largest positive integer $\tiles$ satisfying \eqref{eq: theo: funt, sq, form; proof 3}, is thus the right-hand side rounded down, as given by \eqref{eq: theo: funt, sq, form; funt, simpler}.
\end{proof}

@@

\begin{proof}
% *** Is this writeup too informal/non-rigorous?
% ***LM: I would make the rational-length extension more explicit, and invoke a suitable theorem (with the corresponding book reference) for the real-length case. Or better yet: the first reference linked as a comment in the introduction (https://www.jstor.org/stable/2317945?seq=1#metadata_info_tab_contents) already gives the $2\len/(\pi a)$ formula. We can directly start from that formula, referencing that paper.
The classic Buffon's Needle result (**AA cite Buffon's 1777 paper?** **LM I think maybe something more modern such as the references that are linked in the Introduction section**) states that if a needle of length $1$ is tossed with uniformly random position and orientation on
infinite ruled paper with parallel lines a distance of $1$ apart the probability that it crosses any line is $2/\pi$.

This is also the expected (that is, average) number of crossings, since the needle cannot reach more than one line, ignoring the event of it lying exactly along a line, which has probability $0$. For a needle of length $2$, the expected number of crossings is double the amount of $2/\pi$, since expectation is additive and thus unchanged if the needle is broken into two length-$1$ halves that are tossed separately, and their total crossings are counted. Conversely, for a needle of length $1/2$ the expectation would be half the original one.

A similar argument extends this proportionality to all rational lengths, and by monotonicity all real lengths. Thus for a needle of length $\len$, the expected number of crossings is $2\len/\pi$. If the lines are a distance $a$ apart, this is equivalent to shrinking the needle by a factor of $a$, giving $2\len/(\pi a)$ as expected number of crossings.

Consider a rectangular grid with horizontal spacing $a$ and vertical spacing $b$. Equivalently, this is paper ruled with horizontal lines $b$ apart and vertical lines $a$ apart. The distribution of the number of crossings of a needle tossed onto this grid is known as the Buffon-Laplace Needle Problem (*** cite). This number decomposes into the counts of crossings of horizontal and vertical grid lines, and thus the expected number of crossings is the sum of those for lined paper of width $a$ and $b$ respectively:
\[
\frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right)
\]
As noted in Proposition \ref{prop: i+j-1}, the number of tiles visited by a segment is the count of its grid line crossings plus $1$, unless it exactly crosses any grid points, but again that occurs with probability $0$ and therefore it does not affect the expectation. Thus, $\funta(\len)$ is the above quantity plus $1$.
\end{proof}

In view of Theorem~\ref{theo: funta, form}, the average number of visited tiles as a function of the segment length has a very simple form, namely an affine function. Conversely, for any $\tiles>1$ it is immediate to compute the length of a random segment that visits $\tiles$ tiles on average, given as $\funta^{-1}(\tiles)$.
